---
title: "Understanding protein groups with adjacency matrices"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Understanding protein groups with adjacency matrices}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{PSMatch}
    %\VignetteDepends{mzR,BiocStyle,msdata,SummarizedExperiment}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("PSMatch")`<br />
**Authors**: `r packageDescription("PSMatch")[["Author"]] `<br />
**Last modified:** `r file.info("AdjacencyMatrix.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r setup, message = FALSE, echo = FALSE}
library("PSMatch")
```

# Introduction

This vignette is one among several illustrating how to use the
`PSMatch` package, focusing on the modelling peptide-protein relations
using adjacency matrices and connected componencts. For a general
overview of the package, see the `PSMatch` package manual page
(`?PSMatch`) and references therein.


# Session information

```{r si}
sessionInfo()
```
