---
title: "MS2 fragment ions"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{MS2 fragment ions}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{PSMatch}
    %\VignetteDepends{mzR,mzID,BiocStyle,msdata}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("PSMatch")`<br />
**Authors**: `r packageDescription("PSMatch")[["Author"]] `<br />
**Last modified:** `r file.info("Fragments.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r setup, message = FALSE, echo = FALSE}
library("PSMatch")
```

# Introduction

This vignette is one among several illustrating how to use the
`PSMatch` package, focusing on the calculation and visualisation of
MS2 fragment ions. For a general overview of the package, see the
`PSMatch` package manual page (`?PSMatch`) and references therein.


To illustrate this vignette, we will import and merge raw and
identification data from the `r Biocpkg("msdata")`. For details about
this section, please visit the
[Spectra](https://rformassspectrometry.github.io/Spectra/articles/Spectra.html)
package webpage.

Load the raw MS data:

```{r, message = FALSE}
(spf <- msdata::proteomics(pattern = "2014", full.names = TRUE))
library(Spectra)
sp <- Spectra(spf)
```

Load the identification data:

```{r}
(idf <- msdata::ident(pattern = "2014", full.names = TRUE))
id <- PSM(idf) |> filterPSMs()
id
```

Merge both:

```{r}
sp <- joinSpectraData(sp, id, by.x = "spectrumId", by.y = "spectrumID")
sp
```

In this example, we are going to focus the MS2 scan with index 5449
and its parent MS1 scan (index 5447, selected automatically with the
[filterPrecursorScan()](https://rformassspectrometry.github.io/Spectra/reference/MsBackend.html)
function).

```{r}
sp5449 <- filterPrecursorScan(sp, 5449)
```

```{r}
plotSpectra(sp5449[1], xlim = c(550, 1200))
abline(v = precursorMz(sp5449)[2], col = "red", lty = "dotted")
```

# Calculating fragment ions

The MS2 scan was matched to `SQILQQAGTSVLSQANQVPQTVLSLLR` (there was
obviously no match the the MS1 scan):

```{r}
sp5449$sequence
```

The `calculateFragments()` simply takes a peptide sequence as input
and returns a `data.frame` with the fragment sequences, M/Z, ion type,
charge and position.

```{r}
calculateFragments(sp5449$sequence[2])
```

# Visualising fragment ions

We can now visualise these fragments directly on the MS
spectrum. Let's first visualise the spectrum as is:

```{r}
plotSpectra(sp5449[2])
```

```{r}
plotSpectra(sp5449[2], labels = addFragments, labelPos = 3)
```

# Session information

```{r si}
sessionInfo()
```
