---
title: "Working with PSM data"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Working with PSM data}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{PSMatch}
    %\VignetteDepends{mzR,mzID,BiocStyle,msdata,msdata}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("PSMatch")`<br />
**Authors**: `r packageDescription("PSMatch")[["Author"]] `<br />
**Last modified:** `r file.info("PSMatch.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r setup, message = FALSE, echo = FALSE}
library("PSMatch")
```

# Introduction

This vignette demonstrates how the handle and process proteomics
identification data using the `PSM` package.

# Handling and processing identification data

## Loading PSM data

We are going to use an `mzid` file from the `msdata` package.

```{r}
f <- msdata::ident(full.names = TRUE, pattern = "TMT")
basename(f)
```

The `PSM()` function parses one of multiple `mzid` files and returns
an object of class `DFrame`.


```{r}
library("PSMatch")
id <- PSM(f)
id
```

```{r, echo = FALSE}
n_matches <- nrow(id)
n_scans <- length(unique(id$spectrumID))
n_seqs <- length(unique(id$sequence))
```

This table contains `r n_matches` matches for `r n_scans` scans and
`r n_seqs` peptides sequences.

```{r}
nrow(id) ## number of matches
length(unique(id$spectrumID)) ## number of scans
length(unique(id$sequence))   ## number of peptide sequences
```


The PSM data are read as is, without and filtering. As we can see
below, we still have all the hits from the forward and reverse (decoy)
databases.

```{r}
table(id$isDecoy)
```

## Keeping all matches

The data contains also contains multiple matches for several
spectra. The table below shows the number of number of spectra that
have 1, 2, ... up to 5 matches.

```{r}
table(table(id$spectrumID))
```

Below, we can see how scan 1774 has 4 matches, all to sequence
`RTRYQAEVR`, which itself matches to 4 different proteins:


```{r}
i <- which(id$spectrumID == "controllerType=0 controllerNumber=1 scan=1774")
id[i, ]
```

If the goal is to keep all the matches, but arranged by scan/spectrum,
one can *reduce* the `DataFrame` object by the `spectrumID` variable,
so that each scan correponds to a single row that still stores all
values[^rownames]:

[^rownames]: The rownames aren't needed here are are removed to reduce
    to output in the the next code chunk display parts of `id2`.

```{r, warning = FALSE}
id2 <- reducePSMs(id, id$spectrumID)
rownames(id2) <- NULL ## rownames not needed here
dim(id2)
```

The resulting object contains a single entrie for scan 1774 with
information for the multiple matches stored as lists within the cells.

```{r}
j <- which(id2$spectrumID == "controllerType=0 controllerNumber=1 scan=1774")
id2[j, ]
```

```{r}
id2[j, "DatabaseAccess"]
```

The is the type of complete identification table that could be used to
annotate an raw mass spectrometry `Spectra` object.

## Filtering data

Often, the PSM data is filtered to only retain reliable matches. The
`MSnID` package can be used to set thresholds to attain user-defined
PSM, peptide or protein-level FDRs. Here, we will simply filter out
wrong identification.

### Remove decoy hits

```{r}
id <- filterPsmDecoy(id)
id
```

### Keep first rank matches

```{r}
id <- filterPsmRank(id)
id
```

### Remove non-proteotypic peptides

The peptides below are those that match different proteins. For
example QKTRCATRAFKANKGRAR (the two first rows) matches proteins
`ECA2869` and `ECA4278`.

```{r}
id <- filterPsmNonProteotypic(id)
id
```

Which leaves us with `r nrow(id)` PSMs.

### All filters in one function

This can also be achieved with the `filterPSMs()` function:

```{r}
id <- PSM(f)
filterPSMs(id)
```
# The `mzR` and `mzID` parsers

The `PSM()` function can take two different values for the `parser`
parameter, namely `"mzR"` (which is also the default value) and
`"mzID"`.

- **mzR** uses the `openIDfile()` function from the
   `r BiocStyle::Biocpkg("mzR")` to parse the `mzId` file(s), and then
  coerces the data to a `data.frame` which is eventually returned as a
  `DataFrame`. The parser function uses dedicated code from the
  Proteowizard project (included in `mzR`) and is generally the
  fastest approach.


- **mzID** parses the `mzId` file with `mzID()` function from the
  `r BiocStyle::Biocpkg("mzID")` package, and the flattens the data to
  a `data.frame` with `mzID::flatten()` and eventuelly returns a
  `DataFrame`. The `mzID` package relies on the
  `r BiocStyle::CRANpkg("XML")` package. Is is slower but is is more
  robust to variations in the `mzID` implementation, as is a useful
  backup when the `mzR` approach fails with errors.

```{r, warning = FALSE}
system.time(id1 <- PSM(f, parser = "mzR"))
system.time(id2 <- PSM(f, parser = "mzID"))
```

Other differences in the two parsers include the columns that are
returned, way they name them, and, as will shown below the matches
that are returned. Below, note for instance (and this will be
important later), that there is not equivalent of `"modLocation"` in
`id2`.

```{r}
names(id1)
names(id2)
```

We also have different number of matches in the two tables:

```{r}
nrow(id1)
nrow(id2)
```

```{r}
table(id1$isDecoy)
table(id2$isdecoy)
```

Let's first filter the PSM tables to facilitate focus the comparison
of relevant scans. Note that the default `filterPSMs()` arguments are
set to work with the `mzR` parser.

```{r}
id1_filtered <- filterPSMs(id1)

id2_filtered <- filterPSMs(id2)
```

As can be seen, we are also left with `r nrow(id1_filtered)` vs
`r nrow(id2_filtered)` PSMs after filtering.

The difference doesn't stem from different scans, given that the
spectum identifiers are identical in both tables:

```{r}
identical(sort(unique(id1_filtered$spectrumID)),
          sort(unique(id2_filtered$spectrumid)))
```

The difference is obvious when we tally a table of spectrum ids
occurences in the filtered tables. In `id2_filtered`, each scan is
unique, i.e matched only once.

```{r}
anyDuplicated(id2_filtered$spectrumid)
```

However, for `id1_filtered`, we see that some scans are still repeat
up to 4 times in the table:

```{r}
table(table(id1_filtered$spectrumID))
```

The example below shows that these differences stem from the
modification location (`"modLocation"`), that is not report by the
`mzID` parser:

```{r}
k <- names(which(table(id1_filtered$spectrumID) == 4))
id1_filtered[id1_filtered$spectrumID == k, ]
```

If we remove the `"modLocation"` column, we recoved the same number of
PSMs than with the `mzID` parser.

```{r}
id1_filtered$modLocation <- NULL
nrow(unique(id1_filtered))
```

# Session information

```{r si}
sessionInfo()
```
