---
title: "Validate PSMs based on spectral and identification features"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Validate PSMs}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{PSMatch}
    %\VignetteDepends{BiocStyle,Spectra,PTMods}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("PSMatch")`<br />
**Authors**: `r packageDescription("PSMatch")[["Author"]] `<br />
**Last modified:** `r file.info("validatePSM.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r setup, message = FALSE, echo = FALSE}
library("PSMatch")
library("Spectra")
```

# Introduction {-}

This vignette illustrates how to use the PSM validation functions in the
`PSMatch` package. The validation metrics operate on `Spectra` objects that
contain both raw peak data and peptide identification data, and they can help
assess confidence in peptide–spectrum matches (PSMs) beyond a simple FDR
filter.

The central function `validatePSM()` applies all checks at once and returns a
`data.frame` summarising one row per identified PSM. Each individual check
function can also be called in isolation when only a specific metric is needed.

# Data {-}

## The bundled dataset {-}

This vignette uses `psmBoekweg` and `spBoekweg`, two objects bundled with
`PSMatch`. `psmBoekweg` is a `PSM` object and `spBoekweg` is a `Spectra` object
containing the corresponding raw MS1 and MS2 spectra. More details on these
datasets can be found by loading the `r Biocpkg("MsDataHub")` package and using
`?Boekweg2022`.

```{r load_data}
data("psmBoekweg")
data("spBoekweg")
psmBoekweg
spBoekweg
```

## Merging spectra and identification data {-}

The two objects are joined on a shared key built from the file name and scan
number. It is important to **keep both MS1 and MS2 spectra** in the merged
object as MS1 spectra are required by `checkPrecursorPurity()`.

```{r merge}
psmBoekweg$pkey <- paste0(
    basename(psmBoekweg$filename),
    sub("^.+scan=", "::", psmBoekweg$scannr)
)
spBoekweg$pkey <- paste0(
    basename(spBoekweg$dataOrigin),
    sub("^.+scan=", "::", spBoekweg$spectrumId)
)

sp <- joinSpectraData(spBoekweg, psmBoekweg, by.x = "pkey")
sp
```

## Adding fixed modifications {-}

Depending on the identification data/search engines, the sequence data might
not include modifications such as carbamidomethylation of cysteines. If need be,
we can add fixed modifications by using `PTMods::addFixedModifications()` so
that `calculateFragments()` can account for it when computing fragment m/z
values.

```{r add_mods}
seq_var <- psmVariables(psmBoekweg)[["peptide"]]
fdr_var <- psmVariables(psmBoekweg)[["fdr"]]
## Preview the first few modified sequences
head(sp[[seq_var]][!is.na(sp[[seq_var]])])

## Carbamidomethyl is already present, no need to use:
## sp[["sequence"]] <- addFixedModifications(sp[[seq_var]])

## For converting the annotation style, use `PTMods::convertAnnotation()`
sp[["sequence"]] <- convertAnnotation(sp[[seq_var]], convertToStyle = "name")
head(sp[["sequence"]][!is.na(sp[["sequence"]])])
```

We also create a convenience subset containing only the identified MS2 PSMs
for use with the individual check functions below.

```{r subset_psms}
sp_psms <- sp[!is.na(sp[["sequence"]]) & sp$msLevel == 2L]
```

# Validation metrics {-}

Each section below explains what a metric measures, shows how to call it
directly, and discusses how to interpret the result.

## `checkABpresence()` : a2/b2 ion pair {-}

The a2 and b2 ions are among the smallest b-type and a-type fragments that
can be generated from a peptide. Their simultaneous presence in a spectrum
indicates good fragmentation of an actual peptide ion selected from the MS1.

`checkABpresence()` returns `TRUE` if both the **a2** and **b2** ions are
matched, `FALSE` otherwise. A `FALSE` result does not necessarily mean a bad
identification (it is not abnormal for small ions to be absent), but `TRUE` adds
confidence.

```{r checkAB}
ab <- checkABpresence(sp_psms[1:100])
table(ab)
```

## `checkXYpresence()` : x2/y2 ion pair {-}

Analogous to `checkABpresence()` but for the C-terminal x2/y2 fragment pair.

```{r checkXY}
xy <- checkXYpresence(sp_psms[1:100])
table(xy)
```

## `checkOverlap()` : coverage gap between b- and y-ions {-}

This check evaluates whether the matched b-ions and y-ions together cover the
entire peptide sequence. Given a peptide of length $n$, if the highest matched
b-ion index is $b_max$ and the highest matched y-ion index is $y_max$, a gap
exists when:

$$n - y_max > b_max$$

`checkOverlap()` returns `FALSE` when such a gap is detected and `TRUE` when
b- and y-ions jointly cover the sequence with overlap.

```{r checkOverlap}
## Pre-compute label fragments once to reuse across checks
frags <- labelFragments(sp_psms[1:20], type = c("a", "b", "c", "x", "y", "z"))

overlap <- checkOverlap(sp_psms[1:20],
    peptideVariable = "sequence",
    fragments = frags,
    strippedSeq = PTMods::getCanonicalSequence(sp_psms[["sequence"]][1:20])
)

table(overlap)
```

**Interpretation of a gap.** A gap indicates a region of the peptide that is
not covered by any matched fragment. This often points to an unaccounted-for
modification sitting in the uncovered stretch. Consider the sequence
`sp_psms[10]$sequence`:

```{r overlapPlot}
sp_psms[10]$sequence
checkOverlap(sp_psms[10])
plotSpectraPTM(sp_psms[10])
```
If the identified sequence does not contain a modification yet
`checkOverlap()` returns `FALSE`, applying a modification on one of the
residues in the identified gap could restore fragment coverage.

Conversely, if there are no matched fragments that include the modified
fragments in the sequence, there is no guarantee that there might have been a
modification in the first place.

There is not even a guarantee that those specific amino acids within the gap
are correct or in the correct order.

In short, `checkOverlap()` acts as a ringing bell, indicating poor coverage of
matched fragments within a spectrum and potentially revealing the need to
include or remove a modification.

## `checkShiftConsistency()` : mass shift coverage for modified sequences {-}

When a peptide carries a post-translational modification (PTM), every fragment
ion that includes the modified residue should exhibit the corresponding mass
shift. `checkShiftConsistency()` counts how many of these shift-bearing
fragment positions are actually matched, and returns the ratio as a value
between 0 and 1.

For unmodified sequences the check is not applicable and `NA` is returned.
Higher values mean more of the expected shifted fragments were observed, which
strengthens confidence in both the identification and the modification site
assignment.

```{r checkShift}
shifts <- checkShiftConsistency(sp_psms[1:50])
## NA for unmodified sequences, numeric for modified ones
table(is.na(shifts))
summary(shifts[!is.na(shifts)])
```

Observe that very few fragments including a supposedly identified modification
are matched. The highest value being 0.5833 for sequence `r sp_psms$sequence[3]`
. Out of 12 potential sites including the identified modification, 7 have
been matched:

```{r ex-checkShift}
sp_psms[3]$sequence
checkShiftConsistency(sp_psms[3])
plotSpectraPTM(sp_psms[3])
```

## `checkParentIonIntensity()` : residual precursor in the MS2 spectrum {-}

After CID or HCD fragmentation, the precursor ion is ideally fully fragmented.
If the precursor ion still appears as an intense peak in the MS2 spectrum, it
suggests incomplete fragmentation.

`checkParentIonIntensity()` reports the intensity of the matched precursor
peak relative to the base peak (most intense peak) in the spectrum. A ratio
close to 0 means the precursor was almost fully consumed; a ratio close to 1
means the precursor dominates: a sign of poor fragmentation.

```{r checkParentIon}
parent <- checkParentIonIntensity(sp_psms[1:50])
summary(parent)
```

## `checkPrecursorPurity()` : isolation window contamination {-}

Before fragmentation, the quadrupole isolates a narrow m/z window around the
precursor ion. If other co-eluting peptides happen to fall within this window,
their fragments will contaminate the resulting MS2 spectrum, potentially
leading to chimeric spectra that are harder to interpret.

`checkPrecursorPurity()` wraps `Spectra::precursorPurity()` and computes, for
each MS2 spectrum, the fraction of total signal within the isolation window
(in the preceding MS1 scan) that belongs to the target precursor ion. A purity
of 1 means no contamination; lower values indicate co-isolated species. MS1
spectra and MS2 scans with no preceding MS1 scan return `NA`.

Note that the full `sp` object (MS1 **and** MS2) must be passed here, not just
the PSM subset.

```{r checkPurity}
checkPrecursorPurity(sp[1:6])
```

# Running all checks at once with `validatePSM()` {-}

`validatePSM()` applies all metrics in a single call. It pre-computes
`peaksData` and `labelFragments` internally to avoid redundant computation.
The function operates on a `Spectra` object that contains **both MS1 and
MS2 spectra** alongside the identification data.

```{r validatePSM}
validatePSM(sp[1:400],
    peptideVariable = seq_var,
    fdr = fdr_var
)
```

The returned `data.frame` contains one row per identified PSM with the
following columns:

| Column | Type | Description |
|---|---|---|
| `spectrumId` | character | Spectrum identifier from the raw file |
| `scanIndex` | integer | Scan index within the file |
| `peptide` | character | Modified peptide sequence |
| `canonicalSeq` | character | Sequence without modifications |
| `fdr` | numeric | FDR from the database search |
| `a2b2` | logical | a2 and b2 both matched |
| `x2y2` | logical | x2 and y2 both matched |
| `byOverlap` | logical | Gap detected in b/y coverage |
| `shiftConsistency` | numeric | Fraction of shift-bearing fragments matched (`NA` if no PTM) |
| `parentIonInt` | numeric | Relative precursor intensity in MS2 |
| `precursorPurity` | numeric | Precursor isolation purity from MS1 |

Note the small size of the final dataframe in comparison with the amount of
spectra provided. This shows just how many spectra are not identified by search
engines and the importance to improve the peptide identification in mass
spectrometry-based proteomics.

## Relating metrics back to identification data {-}

The result can be merged with the original PSM table using the spectrum
identifier or the scan index in order to include the metrics for rescoring:

```{r merge_results, eval = FALSE}
## Example: join with the PSM data.frame by spectrum ID
psm_df <- as.data.frame(psmBoekweg)
combined <- merge(psm_df, res,
    by.x = "scannr", by.y = "spectrumId",
    all.x = FALSE
)
```

# Session information {-}

```{r si}
sessionInfo()
```